# Introduction {#intro}


## Introduction to R

- R is a open-source, freely available _statistical programming language_.
- *R Studio* is the most commonly used interface to R.
- Originally developed at Bell Labs in the 1980s as the _S programming language_.
- Modular, most of the functionality is in code libraries called *packages*.. 
- *Literate programming* and *reproducible research* with knitr and markdown.

## The Basics 
- You can think of R as a _really_ fancy calculator, with a _great_ way of handling data.
- The main interaction happens in the *source code window* (by marking code, and pressing Ctrl+R or Apple+R), and on the *R console*.

Addition
```{r, prompt=TRUE}
2+1
```

Subtraction
```{r, prompt=TRUE}
2-1
```

Multiplication
```{r, prompt=TRUE}
2*3
```

Division
```{r, prompt=TRUE}
2/3
```

Exponentiation
```{r, prompt=TRUE}
2^3
```



## Assignment

Values can be stored in **variables**, which have to be named. You might remember the concept from math; the $x$ in $2 = 0.5*x$ is a variable name. 
Variables are created implicitly, by being **assigned** a value. This is done using the R **assignment operator**: `<-`.

For example, let's assign 2 to `var_name`:
```{r, prompt=TRUE}
var_name <- 2
```

Now, let's save the **result** of a **complex expression** ...

```{r, prompt=TRUE}
my_variable <- (11 - 1)^2
my_other_variable <- 10^2
```

... and retrieve it later ...
```{r, prompt=TRUE}
my_variable
```

... or combine with other variables
```{r, prompt=TRUE}
my_variable + my_other_variable
```


## Evaluation 

- Every **expression** **evaluates to** a value. 

- R displays the value an expression evaluates to for almost every **expression** except assignment.

- The result of the addition is shown:

```{r, prompt=TRUE}
1+1
```

- For the assignment, nothing is shown:

```{r, prompt=TRUE}
x <- 1
```


## Operators 

Variables or expressions can be compared with others.
- The operator **'=='** tests for equality: 

```{r, prompt=TRUE}
my_variable == my_other_variable
```

- The operator **'!='** tests for inequality:

```{r, prompt=TRUE}
my_variable != my_other_variable
```


- Don't confuse **'=='** with **'='**. The latter performs assignment (just like **'<-'**):

```{r, prompt=TRUE}
x = 1
```


Variables or expressions can be compared with others.
- What do you think **'<'**, **'>'**, **'<='**, and **'=>'** test for?

```{r, prompt=TRUE}
my_other_variable == 100
```

```{r, prompt=TRUE}
my_other_variable < 100
```

```{r, prompt=TRUE}
my_other_variable <= 100
```


## Using R Help 

- Let's say you forgot the difference between **'='**, **'=='**, and **'<-'**.

```{r, prompt=TRUE, eval=FALSE}
help("<-")
```

```{r, prompt=TRUE, eval=F}
help("=")
```

```{r, prompt=TRUE, eval=FALSE}
help("==")
```

```{r, prompt=TRUE, eval=FALSE}
help("glm")
```


## Data Types

In R, variables have **types**: 

| Data Type  | Examples|
| -----------:|:-------------------------------:|
| Integer (Numeric): | ..., -3, -2, -1, 0, +1, +2, +3, ... |
| Double (Numeric):  | most rational numbers; e.g., 1.0, 1.5, 20.0, pi|
| Character:  | `"a"`, `"b"`, `"word"`, `"hello dear friend, ..."`|
| Logical:  | `TRUE` or `FALSE` (or: `T` or `F` ) |
| Factor:  | Restricted, user-defined set of values, internally represented numerically (e.g., Gender {'male', 'female', 'other'}) |
| Ordered factor:  | Factor with an ordering (e.g., Starbucks coffee sizes {'venti' > 'grande' > 'tall'}) |


```{r, prompt=TRUE}
my_apples <- 1
my_oranges <- "one"
```

Types matter when comparing: 
```{r, prompt=TRUE, error=T}
my_apples == my_oranges
```


Types also matter for other operations.

```{r, prompt=TRUE}
my_apples <- 1
my_oranges <- "1.1"
```
```{r, prompt=TRUE, error=T}
my_apples + my_oranges
```

```{r, prompt=TRUE, error=T}
my_apples + as.double(my_oranges)
```


## Determining Data Types
When unsure, we can ask R about the type of a variable:

```{r, prompt=TRUE}
my_apples <- 1
my_oranges <- "1.1"
```
```{r, prompt=TRUE, error=T}
typeof(my_apples)
```

```{r, prompt=TRUE, error=T}
typeof(my_oranges)
```

```{r, prompt=TRUE, error=T}
typeof(as.double(my_oranges))
```

## Packages
Most of R's functionality is located in packages contributed by users. In this course, we will use the plotting functionality of `ggplot2`.

```{r, prompt=TRUE, eval=FALSE}
# install ggplot2
install.packages("ggplot2")
# install other packages we will use in this course
install.packages(c("tidyverse", "magrittr", "brms"))

# load ggplot2
library(ggplot2)
```


## Plotting

...

## Example: Putting it all (and some more) together

Let's try to understand the Starbucks pricing system.

Here is how we get our data into R.

```{r, prompt=TRUE, error=T}

# store size labels in a vector
label <- c('tall', 'grande', 'venti')
# make it an ordered vector
label <- ordered(label, c('tall', 'grande', 'venti'))

# store sizes in ounces in a vector
size_oz <- c(12, 16, 20)

# store cafe latte prices in a vector; prices probably not up to date
price_tl <- c(8.50, 9.75, 11.00) 

df_starbucks_coffee <- data.frame(label, size_oz, price_tl)
df_starbucks_coffee
```



```{r, prompt=TRUE, error=T}

library(ggplot2)

# plot cup size by size category, use a bar plot;
ggplot(df_starbucks_coffee, aes(label, size_oz)) + geom_bar(stat = "identity")

```

```{r, prompt=TRUE, error=T}

# plot cup size by price, use a scatterplot;  draw a line through the points
ggplot(df_starbucks_coffee, aes(price_tl, size_oz)) + geom_point()  + geom_smooth(method = "lm")

```

```{r, prompt=TRUE, error=T}

# plot cup size by price, use a scatterplot;  draw a line through the points
ggplot(df_starbucks_coffee, aes(price_tl, size_oz)) + geom_point()  + geom_smooth(method = "lm", fullrange=TRUE) + scale_x_continuous(limits = c(5, 15))  + scale_y_continuous(limits = c(0, 25))

```

Does this plot even make sense? Why not?


<!--
You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
-->