
```{r include = FALSE}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed, 
                      prompt = T)
```


```{r result='hide'}

library(ggplot2)
library(magrittr)
library(dplyr)
library(tidyr)

horse_bets_payout_tl <- c(100, -50, 1, 100, -10, -20, 250, -40, -30, 23, -23, 55, 14, 8, 24, -3)
poker_payout_tl <- c(24, 5, -38.1, 12, 103, 15, 5, 187, 13, -23, -45, 36)

horse_bets_profits_tl <- horse_bets_payout_tl - 1.5
poker_profits_tl <- poker_payout_tl

# create two vectors with week days
horse_bets_week_days <- rep(c("Wed", "Fr"), 8)
poker_week_days <- rep(c("Wed", "Fr"), 6)

df_horse_bets <- 
  data.frame(wday = horse_bets_week_days, 
             profit = horse_bets_profits_tl)

df_poker <- 
  data.frame(wday = poker_week_days, 
             profit = poker_payout_tl)

# create a data.frame by concatenating both data.frames
df_gambling <- rbind(df_horse_bets, df_poker)


# let's fix that problem
df_horse_bets$game <- "horse_bets"
df_poker$game <- "poker"
df_gambling <- rbind(df_horse_bets, df_poker)
```


# Working with data

- Today, let's work with realistic data. In order to access it, we'll need to load the R package `languageR`. It provides a number of interesting data sets related to language. 

- You may need to install it, if it's not already installed:

```{r, eval=F, error=T}
install.packages("languageR")
```

- Let's load the package and take a look at the data set.  

```{r, error=T}
library(languageR)

head(dativeSimplified)

```

- This data set happens to be documented. Let's take a look at its description. (You can also see it [here](https://rdrr.io/cran/languageR/man/dativeSimplified.html). )

```{r, eval=F, error=T}
help("dativeSimplified")
```

- The dative alternation in English is the alternation between (1) and (2). Both are grammatical and have the same truth conditions.
- Bresnan et al. (2007) used this data to determine which factors affect the choice between (1) and (2). 

(1) Peter gave Mary the book.
(2) Peter gave the book to Mary.

- Let's take a look at a few rows chosen at random to get a look at a more representative sample ...  

```{r, error=T}
# store all possible row indices in a vector
indices_all <- 1:nrow(dativeSimplified)

# set the random seed to make the results reproducible
set.seed(123)

# choose 10 such numbers at random without replacement
indices_random <- sample(indices_all, size = 10)

# use them to index the data frame to get the corresponding rows
dativeSimplified[indices_random,]
```

- The above code works because rows and columns of data frames can be selected (*'indexed'*) 
  * `df[row_indices, ]` selects rows $-$ ``df[, col_indices]`` selects columns
  * `df[row_indices, col_indices]` selectes rows and columns
  * This works with row and column names, as well as indices.


## Summarizing data

- Let's take a look at the percentages of NP realizations of the recipient by animacy of theme.

```{r, error=T}
# let's take a look at the key dependent variable 
unique(dativeSimplified$RealizationOfRec)

# now, compute the percentages (perc_NP) and the number of observations in each subset
dativeSimplified %>% 
  group_by(AnimacyOfRec) %>% 
  dplyr::summarize(perc_NP = mean(RealizationOfRec == "NP"), 
                   N = n()
                  )
```
### Now, step by step
- Wow, that was easy. ...
- But let's break down the computations into their elementary parts and take look at the logic ...

- First, we'll need to add a column which we can later use to compute the percentage of NP realizations.
```{r, error=T}
# first, let's create a new column encoding whether RealizationOfRec is 'NP'
dativeSimplified$is_realization_NP <- as.integer( dativeSimplified$RealizationOfRec == "NP" )
```

- Take a subset of the data frame where AnimacyOfRec equals "animate" (note the '=='!).
```{r, error=T}
# take a look at the values that can occur in this columns
unique(dativeSimplified$AnimacyOfRec)

# take two subsets corresponding to the values we determined above
df_animate <- subset(dativeSimplified, AnimacyOfRec == "animate")
df_inanimate <- subset(dativeSimplified, AnimacyOfRec == "inanimate")
```

- Compute the average of the column `is_realization_NP` in each data frame, as well as the number of rows in it.
```{r, error=T}
# compute the percentage of NP realizations 
perc_NP_animate = mean(df_animate$is_realization_NP)

# compute the percentage of NP realizations 
N_animate = nrow(df_animate)

# bind to data frame
(res_animate <- data.frame(AnimacyOfRec = "animate", perc_NP = perc_NP_animate, N=N_animate))
```
- Now, do the same for inanimates.
```{r, error=T}
perc_NP_inanimate = mean(df_inanimate$is_realization_NP)
N_inanimate = nrow(df_inanimate)
(res_inanimate <- data.frame(AnimacyOfRec = "inanimate", perc_NP = perc_NP_animate, N=N_animate))
```

- Now, bind the results into a single data frame, which we can use for plotting, creating tables, etc. 

```{r, error=T}
res <- rbind(res_animate, res_inanimate)
res
```

- Puuuuuh. That was tedious.
- The disadvantages of the `subset`-approach are:
  * Values of 'game' need to be known, and so the code needs to be adjusted to the contents of the dataset
  * Much less tractable for more grouping variables
  * Better options are available


## Working with dplyr
- The package *dplyr* ([cheat sheet](http://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf)) provides a set of functions for working with data frames
- Some of functions don't perform a noticeable action, while others reshape data frames or create new ones (the latter are called 'verbs')
```{r, error=T}
library(dplyr)
```

- `arrange()` sorts the data frame by one or several columns
- By default, it does so in ascending order.
```{r, error=T}
arrange(dative, LengthOfTheme)
```
- The modifier `desc()` switches `arrange()` to an decending order. 
```{r, error=T}
arrange(dative, desc(LengthOfTheme))
```

- `select()` keeps *only* the pre-specified columns from the data frame 
```{r, error=T}
# 
select(dative, game, profit)
```

- `left_join` merges two data frames by common columns. This is very useful when data comes from two different sources. 

```{r, error=T}
df1 <- data.frame(name = c("Peter", "Mary", "Susy"), n_apples = c(2,3,4))
df2 <- data.frame(name = c("Peter", "Mary", "Susy"), n_oranges = c(1,20,0))
df <- dplyr::left_join(df1, df2)
df$n_fruit <- df$n_apples + df$n_oranges
df
```

- `sample_n` shows a few random rows from the data frame
```{r, error=T}
sample_n(dativeSimplified, 10)
```

- `group_by` and `ungroup` add and remove information about grouping variables  
```{r, error=T}
# let's 'group' a small subset of the gambling data frame by game
group_by(dativeSimplified[1:2,], RealizationOfRec) 

# let's compare to the original data frame
dativeSimplified[1:2,]
```

- data frames need to be grouped for use with `summarize()` and `mutate()`

```{r, error=T}
# let's group the dativeSimplified data frame 
dativeSimplified_grouped <- group_by(dativeSimplified, game)
```

- `summarize()` divides a grouped data frame into subsets, with *each* subset corresponding to one value of the grouping variable (or a combination of values for several grouping variables), ...
- ... and computes one or several values on each such subset

```{r, error=T}
# summarize several variables
summarize(dativeSimplified_grouped, 
          prop_NP = mean(is_realization_NP),
          prop_animate_rec = mean( AnimacyOfRec == "animate" ),
          prop_animate_theme = mean( AnimacyOfTheme == "animate" ),
          )
```
- `summarize()` can use the results of previous computations
```{r, error=T}
# compute the averages of profit
summarize(dativeSimplified_grouped, 
          prop_NP = mean(is_realization_NP),
          prop_NP = mean( AnimacyOfRec == "animate" ),
          prop_animate_theme = mean( AnimacyOfTheme == "animate" ),
          )
```
- The last line uses the function `ifelse` 


- `mutate()` proceeds similarly to `summarize()` in dividing a grouped dataset into subsets, but instead of computing one or several values for each subset, it creates or modifies a column 

- A very simple application is to simply create a new column. In this case, we don't even need to group
```{r, error=T}
# these two lines performs exactly the same action, except the latter stores the result in df 
dativeSimplified$is_realization_NP <- (dativeSimplified$RealizationOfRec == "NP" ) 
df <- mutate(dativeSimplified, is_realization_NP = (RealizationOfRec == "NP") )

head(df, 2)
head(dativeSimplified, 2)
```


## Pipes

### The problem

- The code below is really hard to read, even harder to maintain, and `dativeSimplified_grouped_by_AnimacyOfRec_and_AnimacyOfTheme` is a terribly long variable name.
```{r, error=T}
dativeSimplified_grouped_by_AnimacyOfRec_and_AnimacyOfTheme <-
      group_by(dativeSimplified, AnimacyOfRec, AnimacyOfTheme)
df <- mutate(dativeSimplified_grouped_by_AnimacyOfRec_and_AnimacyOfTheme, 
                  perc_NP = mean(RealizationOfRec == "NP") )
df
```

- This alternative is also quite bad. To read this code, you need to know which bracket matches which other bracket.
```{r, error=T}
df <- mutate(group_by(dativeSimplified, AnimacyOfRec, AnimacyOfTheme), 
                  perc_NP = mean(RealizationOfRec == "NP") )
df
```

- One nested function call may be OK. But try to read this.
```{r, error=T}
df <- dplyr::summarize(group_by(mutate(dativeSimplified, long_theme = ifelse(LengthOfTheme > 1.6, "long", "short") ), long_theme),
              perc_NP = mean(RealizationOfRec == "NP")
              )
```

- Or consider this expression (`sqrt` is the square root.)
```{r, error=T}
sqrt(divide_by(sum(divide_by(2,3), multiply_by(2,3)), sum(3,4)))
```

- Luckily, there a better way to write this expression.

### Pipes
- The problem is that we have too many levels of embedding.
- In natural language we avoid multiple embeddings of that sort by making shorter sentences, and using anaphors to refer to previous discourse.
- The packages **dplyr** and **magrittr** provide a limited version of such functionality, and we'll need to use **pipe** operators (`%>%` and `%<>%`) to link expressions with an 'anaphoric dependency'.

```{r, error=T}
# Typical notation. Read as "Divide 10 by 2."
divide_by(10, 2)

# Equivalent pipe notation. Read as "Take 10, and divide it by 2."
10 %>% divide_by(., 2)

# Equivalent pipe notation. Read as "Take 2, and divide 10 by it."
2 %>% divide_by(10, .)
```

- If the dot operator occurs in the first argument slot, it can be omitted. (R has pro-drop.)
```{r, error=T}
# pipe notation with omission of '.'
10 %>% divide_by(2)
```

- Let's see how it can resolve the mess below. (Repetition of previous example.)
```{r, error=T}
df <- mutate(group_by(dativeSimplified, AnimacyOfRec, AnimacyOfTheme), 
                  perc_NP = mean(RealizationOfRec == "NP") )
df
```

- And here is the much more readable version of this code:
```{r, error=T}
df <-  dativeSimplified %>%
          mutate(., long_theme = ifelse(LengthOfTheme > 1.6, "long", "short") ) %>% 
          group_by(., long_theme) %>% 
          dplyr::summarize(., perc_NP = mean(RealizationOfRec == "NP") )
```
- We don't actually need the dot:
```{r, error=T}
df <-  dativeSimplified %>%
          mutate(long_theme = ifelse(LengthOfTheme > 1.6, "long", "short") ) %>% 
          group_by(long_theme) %>% 
          dplyr::summarize(perc_NP = mean(RealizationOfRec == "NP") )
```


- The `%<>%` operator is a convenient combination of `%>%` and `<-` which you can use to directly modify an object.

```{r, error=T}
# create a vector with numbers from 1 to 10
x <- 1:10
# keep only numbers < 5:
#   (i) without %<>%
x <- x[x<5]
#   (i) with %<>%
x %<>% .[.<5]
```

```{r, error=T}
# lets add several columns to 'dativeSimplified'
dativeSimplified %<>% mutate(A=1, B=2, C=3, D=4)
dativeSimplified
```

