--- 
title: "Data Processing in R, Lecture Notes"
author: "Pavel Logacev"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Lecture notes on data processing in R."
---

<!--
Markdown basics: https://rmarkdown.rstudio.com/authoring_basics.html
-->

# Getting Started

## Initial Remarks

- The purpose of these lecture notes is to allow you to quickly revisit concepts covered in class. On their own, they likely won't form a sufficient introduction to R.
- In the following I will draw on the following resources:
  - The great introduction materials developed at the University of Glasgow: <https://psyteachr.github.io/>, in particular ['Data Skills for Reproducible Science'](https://psyteachr.github.io/msc-data-skills/).
  - Primers on a variety of topics: https://rstudio.cloud/learn/primers
  - Cheat sheets on a variety of topics: https://rstudio.cloud/learn/cheat-sheets
  
<!--
- the O'Reiley book
- 
-->

## Installing and Setting up RStudio
Currently, the most convenient interface to R is the RStudio GUI. You can either install it locally, or use it online on RSudio Cloud. [RStudio Cloud]().

### Local RStudio installation
- Typically, *RStudio* is run on a local computer. 

Please download R and RStudio separately from:
  - https://cran.r-project.org/
  - https://rstudio.com/products/rstudio/download/ (download 'RStudio Desktop')
  
### RStudio Cloud
You can use the RStudio cloud service for free after getting an account to https://rstudio.cloud.

<!--
## RStudio Interface 
- Let's take a look at the RStudio interface.
- Let's take a look at the [RStudio Cloud](https://rstudio.cloud) interface.
-->

## Basics of the RStudio environment
Please take a look at the following [thorough introduction to the RStudio environment](https://github.com/hbctraining/Intro-to-R/blob/master/lessons/01_introR-R-and-RStudio.md).

Please watch the following video by Lisa DeBruine:
- ['RStudio Setup'](https://www.youtube.com/watch?v=CbA6ZVlJE78&list=PLA2iRWVwbpTIweEBHD2dOKjZHK1atRmXt&index=2&ab_channel=LisaDeBruine)

<!--
This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.org/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
-->